---
title: "Introduction to LargeScaleCharacteristic"
author: "Marti Catala, Yuchen Guo"
date: '2023-02-17'
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Introduction to LargeScaleCharacteristic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(LargeScaleCharacteristics)
```


## Introduction
In this vignette we show the two functions in this package to get characteristics and/or covariates/features. The two functions in this package are:

1)`getLargeScaleCharacteristics()` - this function will provide a summary of characteristics for your original cohort in tables interested, with regard to target windows.

2)`addLargeScaleCharacteristics()` - this function will add extra columns to your original cohort, with covariates/features from tables interested.

By tables interested, in this package we support the following OMOP Common Data Model (CDM) tables:
  "visit_occurrence", "condition_occurrence", "drug_exposure", "procedure_occurrence", "device_exposure", "measurement", "observation", "drug_era", "condition_era" and "specimen"




## Get summary characteristics using getLargeScaleCharacteristics
Now we first show an example using getLargeScaleCharacteristics(). As LargeScaleCharacteristics package works with data mapped to the OMOP CDM and cdm object, we will first need to connect to a database, and use the CDMConnector package to represent our mapped data as a single cdm object in R. This could like something like:
```{r eval=FALSE}
con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = Sys.getenv("CDM5_POSTGRESQL_DBNAME"),
                      host = Sys.getenv("CDM5_POSTGRESQL_HOST"),
                      user = Sys.getenv("CDM5_POSTGRESQL_USER"),
                      password = Sys.getenv("CDM5_POSTGRESQL_PASSWORD"))
cdm <- CDMConnector::cdm_from_con(con,
                                  cdm_schema = Sys.getenv("CDM5_POSTGRESQL_CDM_SCHEMA"))
```
For this example we´ll generate mock cdm data using `mockLargeScaleCharacteristics()` function. We first create a cdm object containing cohort1 (target cohort) and drug_exposure table (table to characterise)
```r
cohort1 <- tibble::tibble(
  cohort_definition_id = c("1", "1"),
  subject_id = c("1", "2"),
  cohort_start_date = c(
    as.Date("2010-04-01"),
    as.Date("2010-04-01")
  ),
  cohort_end_date = c(
    as.Date("2011-06-01"),
    as.Date("2011-06-01")
  )
)

drug_exposure <- tibble::tibble(
  person_id = c("1", "2"),
  drug_exposure_start_date = c(
    as.Date("2009-05-01"),
    as.Date("2009-05-01")
  ),
  drug_exposure_end_date = c(
    as.Date("2010-03-03"),
    as.Date("2010-03-03")
  ),
  drug_concept_id = c("1", "1")
)

cdm <- mockLargeScaleCharacteristics(
  cohort1 = cohort1,
  drug_exposure = drug_exposure
)
```

The cdm object have tables look like this now:

```r
cdm$drug_exposure
 # Source:   table<drug_exposure> [2 x 4]
 # Database: DuckDB 0.5.0 [root@Darwin 21.3.0:R 4.2.1/:memory:]
   person_id drug_exposure_start_date drug_exposure_end_date drug_concept_id
   <chr>     <date>                   <date>                 <chr>          
 1 1         2009-05-01               2010-03-03             1              
 2 2         2009-05-01               2010-03-03             1   

cdm$cohort1
 # Source:   table<cohort1> [2 x 4]
 # Database: DuckDB 0.5.0 [root@Darwin 21.3.0:R 4.2.1/:memory:]
   cohort_definition_id subject_id cohort_start_date cohort_end_date
   <chr>                <chr>      <date>            <date>         
 1 1                    1          2010-04-01        2011-06-01     
 2 1                    2          2010-04-01        2011-06-01    
```
Get large scale characteristic using function getLargeScaleCharacteristics() with output a tiblble with following columns: cohort_definition_id, table_id (in this case 1 as only characteristics from drug_exposure table are requested), window_id, window_name, concept_id, concept_name, concept_count, denominator_count, overlap and concept_type. Here we set minimumCellCount to 1 first, we demonstrate the count is the correct number we want. For this example, both person 1 & 2 has drug exposure record in windows: c(-365;-91), c(365;-31), c(90;-1), c(-30;-1). So we expect all concept_count and denominator_count to be 2. Below we run the getLargeScaleCharacteristics() function
\tiny
``` r
getLargeScaleCharacteristics(cdm,
                             targetCohortName = c("cohort1"),
                             overlap = TRUE,
                             tablesToCharacterize = "drug_exposure",
                             minimumCellCount = 1
)
#> # this is what the function output looks like:
#> # A tibble: 4 × 11
#>   cohort_definition_id table_id table_name    window_id window_name concept_id concept_name   concept_count denominator_count overlap concept_type
#>   <chr>                   <int> <chr>             <int> <chr>       <chr>      <chr>          <chr>         <chr>             <lgl>   <chr>       
#> 1 1                           1 drug_exposure         2 -365;-91    1          concept_name_1 2             2                 TRUE    Standard    
#> 2 1                           1 drug_exposure         3 -365;-31    1          concept_name_1 2             2                 TRUE    Standard    
#> 3 1                           1 drug_exposure         4 -90;-1      1          concept_name_1 2             2                 TRUE    Standard    
#> 4 1                           1 drug_exposure         5 -30;-1      1          concept_name_1 2             2                 TRUE    Standard    
```
\normalsize
minimumCellCount is a variable you can set, and all counts lower than minimumCellCount will be output as '< minimumCellCount'. Here we run the same function but reset the minimumCellCount to 3. (With default this value is set to 5)
\tiny
``` r
getLargeScaleCharacteristics(cdm,
                             targetCohortName = c("cohort1"),
                             overlap = TRUE,
                             tablesToCharacterize = "drug_exposure",
                             minimumCellCount = 3
)
#> # now the output looks like this:
#>  A tibble: 4 × 11
#>   cohort_definition_id table_id table_name    window_id window_name concept_id concept_name   concept_count denominator_count overlap concept_type
#>   <chr>                   <int> <chr>             <int> <chr>       <chr>      <chr>          <chr>         <chr>             <lgl>   <chr>       
#> 1 1                           1 drug_exposure         2 -365;-91    1          concept_name_1 <3            <3                TRUE    Standard    
#> 2 1                           1 drug_exposure         3 -365;-31    1          concept_name_1 <3            <3                TRUE    Standard    
#> 3 1                           1 drug_exposure         4 -90;-1      1          concept_name_1 <3            <3                TRUE    Standard    
#> 4 1                           1 drug_exposure         5 -30;-1      1          concept_name_1 <3            <3                TRUE    Standard     
```
\normalsize
## Add characteristics columns using addLargeScaleCharacteristics

Package also allows you to add column to a cohort using table(s) interested in, with function addLargeScaleCharacteristics(). Here we show an example of add covariates from drug exposure table to cohort1. First we define a drug exposure table with 5 different drug_concept_id:

\tiny
``` r
drug_exposure <- tibble::tibble(
  person_id = c("1", "1", "1", "1", "1"),
  drug_exposure_start_date = c(
    as.Date("2009-05-01"),
    as.Date("2009-05-01"),
    as.Date("2009-05-01"),
    as.Date("2009-05-01"),
    as.Date("2009-05-01")
  ),
  drug_exposure_end_date = c(
    as.Date("2010-03-03"),
    as.Date("2010-03-03"),
    as.Date("2010-03-03"),
    as.Date("2010-03-03"),
    as.Date("2010-03-03")
  ),
  drug_concept_id = c("1", "2", "3", "4", "5")
)

cdm <- mockLargeScaleCharacteristics(
  cohort1 = cohort1,
  drug_exposure = drug_exposure
)

cdm$drug_exposure %>% dplyr::select(drug_concept_id) %>% dplyr::distinct()
#> # Source:   SQL [5 x 1]
#> # Database: DuckDB 0.5.0 [root@Darwin 21.3.0:R 4.2.1/:memory:]
#>   drug_concept_id
#>             <dbl>
#> 1               1
#> 2               5
#> 3               2
#> 4               4
#> 5               3
```
\normalsize
Hence, the columns we are adding to cohort will have the name form of: {table_name}_{concept_id}_{window_name}. To demonstrate, we only select one window in each run separatemly, this example list(c(-30, -1)) and list(c(91, 365)). The result for window list(c(-30, -1)) looks like this. (NAs are generated as subject 2 is not in tablesToCharacterize (drug_exposure))
\tiny
``` r
addLargeScaleCharacteristics(
  x = cdm$cohort1,
  cdm,
  overlap = TRUE,
  temporalWindows = list(c(-30, -1)),
  tablesToCharacterize = c("drug_exposure")
)
#> The result for window list(c(-30, -1)) is like this: 
#> # Source:   table<dbplyr_595> [2 x 8]
#> # Database: DuckDB 0.5.0 [root@Darwin 21.3.0:R 4.2.1/:memory:]
#>   subject_id cohort_start_date cohort_end_date `drug_exposure_5_-30;-1` `drug_exposure_4_-30;-1` `drug_exposure_3_…` `drug_exposure…` `drug_exposure…`
#>   <chr>      <date>            <date>                             <dbl>                    <dbl>               <dbl>            <dbl>            <dbl>
#> 1 1          2010-04-01        2011-06-01                             1                        1                   1                1                1
#> 2 2          2010-04-01        2011-06-01                            NA                       NA                  NA               NA               NA
```
\normalsize
Now the result for window list(c(91, 365)) looks like this. NAs are generated as subject 2 is not in tablesToCharacterize (drug_exposure). 0s are generated as subject 1 does not have drug exposure events fall in this window list(c(91, 365))
\tiny
```r
addLargeScaleCharacteristics(
  x = cdm$cohort1,
  cdm,
  overlap = TRUE,
  temporalWindows = list(c(91, 365)),
  tablesToCharacterize = c("drug_exposure")
)
#> Now result of window list(c(91, 365)) is like this: 
#> # Source:   table<dbplyr_601> [2 x 8]
#> # Database: DuckDB 0.5.0 [root@Darwin 21.3.0:R 4.2.1/:memory:]
#>   subject_id cohort_start_date cohort_end_date `drug_exposure_5_91;365` `drug_exposure_4_91;365` `drug_exposure_3_…` `drug_exposure…` `drug_exposure…`
#>   <chr>      <date>            <date>                             <dbl>                    <dbl>               <dbl>            <dbl>            <dbl>
#> 1 1          2010-04-01        2011-06-01                             0                        0                   0                0                0
#> 2 2          2010-04-01        2011-06-01                            NA                       NA                  NA               NA               NA

```



